---
title: "Omega PLC"
author: "Shivant Maharaj"
date: "2022-09-19"
output:   
html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)

```

# Introduction

The following assessment was given as a classroom exercise to test our
understanding of R, Version Control, and Visualisations using GGPlot2.

# Omega Group plc- Pay Discrimination

At the last board meeting of Omega Group Plc., the headquarters of a
large multinational company, the issue was raised that women were being
discriminated in the company, in the sense that the salaries were not
the same for male and female executives.

We were asked to carry out the analysis to find out whether there is a
significant difference between the salaries of men and women, and if
there are any discrimination factors.

## Loading the data

Omega Group Plc. had shared data of 50 employees with us. The data was
inspected before conducting our analysis and we have noted that there
were no missing/incomplete data points.

```{r load_omega_data, warnings = FALSE}
omega <- read_csv(here::here("data", "omega.csv"))
skimr::skim(omega) # examine the data frame
```

## Relationship Salary - Gender ?

The data frame `omega` contains the salaries for the sample of 50
executives in the company.

We have constructed a visualization that shows the salary trend between
male and female employees:

```{r, salary_visualisation}

#Extra Visual
ggplot(omega) + 
  geom_boxplot() +
  aes(x = salary, y = gender, colour = gender) + 
  labs (x = "Salary", y = "Gender", caption = "Source: Omega Plc salary data", title = "Male employees at Omega Plc have a higher salary than females", subtitle = "Boxplot showing the salary trend between males and females at Omega Plc")
```

The above graph shows a visual difference between the average
distributions of salary between female and male salary's for Omega Plc.
The group decided to extrapolate the data and construct tables depict
these initial findings.

```{r, salary_table}
# Summary Statistics of salary by gender
mosaic::favstats (salary ~ gender, data=omega)

#Summary Stats 

summary_stats <- omega %>% 
  group_by(gender) %>% 
  summarise(mean_salary = mean(salary),
            sd_salary = sd(salary),
            count = n(),
            t_critical = qt(0.975, count-1),
            se_salary = sd(salary)/sqrt(count),
            margin_of_error = t_critical * se_salary,
            salary_low = mean_salary - margin_of_error,
            salary_high = mean_salary + margin_of_error)

summary_stats

obs_diff <- omega %>%
  specify(salary ~ gender) %>%
  calculate(stat = "diff in means", order = c("male", "female"))
# Dataframe with two rows (male-female) and having as columns gender, mean, SD, sample size, 
# the t-critical value, the standard error, the margin of error, 
# and the low/high endpoints of a 95% condifence interval

p_val <- summary_stats %>%
  get_p_value(obs_stat = obs_diff, direction = "two_sided")

p_val

```

From the initial analysis in the table above, it is shown that males
have a higher mean salary of 8696 than females. We decided that more
analysis is required to determine if this difference is significant and
to decide if there are any further analysis required to uncover
influencing factors.

### Advanced Analysis

The team decided to conduct 2 tests to decide if the difference in mean
salary's between males and females were significant. The following tests
were decided upon:

1.  Hypothesis Tesing useing the T Test package
2.  T Test using simulation method and infer packages

#### Hypothesis Tesing using the T Test package

The T Test package was used to assess if there is a true difference in
the salaries, and findings form this built-in test shows that there is
indeed a significant difference as indicated by the absolute T-Value
being greater than 1.96, and the P-Value being smaller than 5%.

```{r, hypothesis_testing_ttest}
# hypothesis testing using t.test() 
t.test(salary ~ gender, data = omega)

```

We also ran a hypothesis test, assuming that the mean difference in
salaries is zero as the null hypothesis using the simulation method from
the `infer` package.

```{r, hypothesis_testing_infer}
# hypothesis testing using infer package
library(infer)
set.seed(1234)
infer_stats <- omega %>% 
  specify(salary ~ gender) %>% 
  hypothesize(null = "independence",) %>% 
  generate(reps = 1000, type ="permute") %>% 
  calculate(stat = "diff in means", order = c("male", "female"))

percentile_ci <- infer_stats %>% get_confidence_interval(level = 0.95, type = "percentile")

visualize(infer_stats) + shade_p_value(obs_stat = obs_diff, direction = "two-sided")
percentile_ci


```

From both tests conducted, we can conclude that the observed difference
in the means of salaries between males and females at Omega Plc is
indeed a significant difference.

As depicted in the simulation density visualisation above, we can see
that the observed difference is passed the Upper 95% percentile of 4829.
This is confirmed by the t.test() performed that states the absolute
T-value as 4, which is much bigger than the standard acceptable value of
1.96. Additionally, the p-value was smaller than 1%.

## Relationship Experience - Gender?

At the board meeting, someone raised the issue that there was indeed a
substantial difference between male and female salaries, but that this
was attributable to other reasons such as differences in experience. A
questionnaire send out to the 50 executives in the sample reveals that
the average experience of the men is approximately 21 years, whereas the
women only have about 7 years experience on average (see table below).

```{r, experience_stats}
# Summary Statistics of experience by gender
favstats (experience ~ gender, data=omega)

#Summary Stats
summary_stats_exp <- omega %>% 
  group_by(gender) %>% 
  summarise(mean_experience = mean(experience),
            sd_experience = sd(experience),
            count = n(),
            t_critical = qt(0.975, count-1),
            se_experience = sd(experience)/sqrt(count),
            margin_of_error = t_critical * se_experience,
            experience_low = mean_experience - margin_of_error,
            experience_high = mean_experience + margin_of_error)

summary_stats_exp


```

The above data shows a difference in experience mean of 13.74 years
between females and males. To establish if the observation is a
significant difference, further analysis is required.

```{r , t test experience}
t.test(experience ~ gender, data = omega)

```

Further analysis was conducted in the form of a T Test to assess the if
there is any significant differences in experience between genders. The
following test show the statistical variables for Omega Plc.The findings
show that there is a significant difference in the experience between
males and females from Omega Plc, shown by the observed t-value is
greater than 1.96 and a small p-value. This finding assists in
validating the pervious observation - as the average experience of males
suggests reason as to why the salaries of genders are different.

Based on this evidence, can you conclude that there is a significant
difference between the experience of the male and female executives?
Perform similar analyses as in the previous section. Does your
conclusion validate or endanger your conclusion about the difference in
male and female salaries?

## Relationship Salary - Experience ?

To further substantiate the findings from the previous tests conducted,
a final visual check is performed to determine if there have been any
discrimination against females at Omega Plc.

```{r, salary_exp_scatter}
ggplot(omega)+
  aes(x = experience, y = salary, color = gender)+
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Experience", y = "Salary", title = "Female's yearly increases are greater than that of men", subtitle = "Scatterplot showing the distribution of salary versus experience for males and females.")


```

The plot shows that there are a greater proportion of female employees
at Omega Plc with less than 10 years of experience as compared to males,
while there is a greater proportion of males with more than 10 years of
experience than females at the company

## Correlation checks between gender, experience, and salary

The following graph shows a visual matrix of how gender and experience
affect salary.

```{r, ggpairs}
omega %>% 
  select(gender, experience, salary) %>% #order variables they will appear in ggpairs()
  ggpairs(aes(colour=gender, alpha = 0.3))+
  theme_bw()
```

The visual assessing salary and experience for various genders shows the
same finding as described in the visual before.
