---
title: "Weather Analysis in R"
author: "Shivant Maharaj"
date: "2022-09-15"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(wbstats)
library(countrycode)
library(patchwork)
library(gganimate)
library(scales)
library(infer)
library(rayshader)
library(gganimate)
```

# Climate change and temperature anomalies

We are analysing a dataset from NASA's Goddard Institute for Space
Studies to study the effects of climate change in the Northern
Hemisphere. Glimpsing at the data, we see there are 19 variables and 143
observations, representing the period between 1880-2022:

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")
glimpse(weather)

```

For the purpose of our analysis, we have decided to select only data
pertaining to temperature deviation (delta) by month, and manipulate the
dataframe to facilitate further investigation:

```{r tidyweather}

tidyweather <- select(weather, 1:13) %>% 
  pivot_longer(!Year, names_to = 'month', values_to = 'delta')
head(tidyweather)
  

```

## Plotting Information

First, we are plotting a scatter plot to visualize the evolution of
delta (temperature deviation) over time:

```{r scatter_plot}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

gg <- ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Weather Anomalies"
  )

#plot_gg(gg,multicore=TRUE,width=5,height=5,scale=250)

```

Adding a line of best fit to the scatterplot clearly shows that, while
deltas were close to 0 between approximately 1935-1975 and negative
before that time, temperature in the Northern Hempishere has been
quickly increasing since then. Notice that the rate of the increase has
been increasing as well (as signified by increasing deltas).

Next, we will use `facet_wrap()` to visualize the effects of increasing
temperature by month:

```{r facet_wrap, echo=FALSE}

ggplot(tidyweather, aes(x= year, y = delta))+
  geom_point()+
  geom_smooth(color="red") +
  facet_wrap(~month)+
  theme_bw() +
  labs (
    title = "Weather Anomalies"
  )

```

We can see that the effect of rising temperature in the Northern
Hemisphere is common to all months of the year.

As a means to further investigate the effects of climate change, we will
partition the data into time periods, particularly decades. To that end,
we will use `case_when()`:

```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```

In order to study the effects of climate change by decade, we will
produce a density plot to investigate the distribution of monthly
temperature deviations by the time periods selected above:

```{r density_plot }

ggplot(comparison) +
  aes(x = delta, fill = interval)+
  #facet_wrap(~month)+
  geom_density(alpha = 0.2) 

```

The plot clearly shows that with the passage of time, deltas have
consistently moved to the right hand side of the plot. In other words,
temperature deviations have been increasing over time.

Lastly, we will also consider annual anomalies by grouping the monthly
data and producing a scatterplot:

```{r averaging }

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(yearly_mean = mean(delta, na.rm=TRUE)) 
  
average_annual_anomaly
#plotting the data
#Fit the best fit line, using LOESS method
ggplot(average_annual_anomaly) +
  aes(x = Year, y = yearly_mean)+
  geom_point()+
  geom_smooth(method = 'lm') +
  theme_bw()
```

The plot proves that annual temprature deltas have been increasing over
time.

## Confidence Interval for `delta`

We will now focus on the time period between 2011-present. We divert our
attention towards producing a confidence interval for the average annual
deltas calculated since 2011. We will use both the statistical method
and bootstrap simulation to have more confidence in the results:

```{r, calculate_CI_using_stats }

#statistical method
formula_ci <- comparison %>% 
  filter(interval == '2011-present') %>% 
  group_by(year) %>% 
  summarise(avg_annual_delta = mean(delta),
            sd_delta = sd(delta),
            count = n(),
            SE = sd(delta/sqrt(count)),
            upper_ci = avg_annual_delta + 2*SE,
            lower_ci = avg_annual_delta - 2*SE)

#print out formula_CI
formula_ci
```


```{r , bootstrap}
#bootstrap simulation  

set.seed(1234)
bootstrap <- comparison %>% 
  filter(interval == '2011-present') %>% 
  specify(response = delta) %>% 
  generate(type = 'bootstrap') %>% 
  calculate(stat = 'mean') 

bootstrap_ci <- bootstrap %>% get_confidence_interval(level = 0.95)

#print out confidence intervals
bootstrap_ci

#visualise
visualise(bootstrap)

```


```{r}
library(gapminder)
library(gganimate)
library(png)
library(gifski)

#skimr::skim(gapminder)

# Use the gapminder dataset in ggplot
ggplot(data=gapminder,
       aes(x=gdpPercap, y=lifeExp, size=pop, color=country)) +
  # Add a point geom
 geom_point(alpha=0.7, show.legend=FALSE) +
# Add some manual scaling and facets 
 scale_colour_manual(values=country_colors) +
 scale_size(range=c(2, 12)) +
 scale_x_log10() +
 facet_wrap(~continent, nrow=1) + 
# Animate figure with gganimate package
 transition_time(year) +
 ease_aes('linear') +
 labs(title='Year: {frame_time}', 
      x='GDP per capita', 
      y='Life expectancy')
```

Looking at the results of the analysis, we can see that the statistical
method produces wider confidence intervals for temperature deltas,
ranging from 0.13 to approximately 0.3 in width. This is probably due to
the low number of observations (12 months in each year), which prohibit
a more precise calculation. On the other hand, using bootstrap
simulation allows to get a much narrower confidence interval. However,
both methods show that temperature deltas have been positive in the time
period in question and have been consistently greater than 1 since 2015.

